---
title: "03 - Biomass recovery without spatial recovery"
subtitle: "Results 2.1: change-point, hysteresis, matched-biomass contrast, and the survey-rescaling sensitivity (Supplementary Table S1)"
author: "Sarah Ohayon, Jacob T. Dingwall, Amanda E. Bates"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")
source("00_setup.R")
library(strucchange); library(knitr); library(mgcv); library(patchwork)
library(readxl); library(scales)
```

Every statistic reported in Results §2.1 is produced here. The argument has four steps:

1. the occupancy series contains an abrupt structural break, independent of any phase we impose;
2. once that break is accounted for, biomass has no detectable effect on occupancy;
3. contrasting the two phases *at equivalent biomass* shows the same gap;
4. the conclusion survives any plausible rescaling of the older surface-survey index
   (**Supplementary Table S1**), including corrections large enough to erase the apparent
   biomass recovery altogether.

```{r load}
annual    <- read_csv(file.path(DERIVED, "annual_series.csv"), show_col_types = FALSE)
site_year <- read_csv(file.path(DERIVED, "site_year_spawn.csv"), show_col_types = FALSE)

d <- annual %>%
  select(Year, n_active, occupancy, SSB) %>%
  filter(!is.na(occupancy)) %>%
  mutate(SSBk  = SSB/1000,
         phase = factor(if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"),
                        levels = c("Pre-collapse","Post-collapse"))) %>%
  arrange(Year)

N_all <- round(unique(d$n_active / d$occupancy)[1])
cat(sprintf("Occupancy denominator N_all = %d sections; series %d-%d (n = %d years)\n",
            N_all, min(d$Year), max(d$Year), nrow(d)))
```

---

## 1  The occupancy series contains a structural break

A supF test asks whether *any* break is present; the BIC-optimal single-break model then
locates it and returns a 95% confidence interval. Neither uses the 1984 boundary as an
input, so the break year is estimated, not assumed.

```{r changepoint}
sc <- sctest(d$occupancy ~ 1, type = "supF")
print(sc)

bp <- breakpoints(d$occupancy ~ 1, breaks = 1)
ci <- confint(bp, breaks = 1)$confint

cat(sprintf("\nBreak after %d (new regime from %d); 95%% CI %d-%d\n",
            d$Year[bp$breakpoints], d$Year[bp$breakpoints + 1],
            d$Year[ci[1]], d$Year[ci[3]]))
cat(sprintf("supF = %.0f, P %s\n", sc$statistic, fmt_p(sc$p.value)))

cat(sprintf("\nLargest single-year decline: %d -> %d, %d -> %d occupied sections\n",
            1983, 1984, d$n_active[d$Year == 1983], d$n_active[d$Year == 1984]))

pre  <- d %>% filter(phase == "Pre-collapse")
post <- d %>% filter(phase == "Post-collapse")
cat(sprintf("Pre-collapse  (<=%d): %.1f +/- %.1f sections (%.0f%%)\n",
            BREAK_YEAR-1, mean(pre$n_active), sd(pre$n_active), 100*mean(pre$occupancy)))
cat(sprintf("Post-collapse (>=%d): %.1f +/- %.1f sections (%.0f%%)\n",
            BREAK_YEAR, mean(post$n_active), sd(post$n_active), 100*mean(post$occupancy)))
cat(sprintf("Decline in occupied sections: %.0f%%\n",
            100*(mean(pre$n_active) - mean(post$n_active))/mean(pre$n_active)))
```

### The contraction precedes the change in survey method

Dive surveys began in the Strait of Georgia in 1985 and immediately dominated the index. The contraction occurred in 1984, while
every spawn record in the Strait was still a surface survey - so it cannot be an artefact of
the method transition.

```{r survey-method}
spawn_raw <- read_csv(F_SPAWN, show_col_types = FALSE) %>%
  mutate(StatisticalArea = suppressWarnings(as.numeric(StatisticalArea))) %>%
  keep_surveyed()   # presence rule, 00_setup.R
meth <- spawn_raw %>% filter(Region == SOG_REGION, StatisticalArea %in% OCC_AREAS, !is.na(Year)) %>%
  mutate(period = if_else(Year <= 1984, "1951-1984", "1985-2025"),
         bio_surface   = replace_na(Surface, 0),
         bio_under     = replace_na(Understory, 0),
         bio_macro     = replace_na(Macrocystis, 0))
meth %>% group_by(period) %>%
  summarise(n_records = n(),
            pct_surface_derived   = 100*sum(bio_surface)/sum(bio_surface+bio_under+bio_macro),
            pct_understory_derived= 100*sum(bio_under)  /sum(bio_surface+bio_under+bio_macro),
            .groups = "drop") %>%
  mutate(across(where(is.numeric), ~round(., 1))) %>%
  kable(caption = "Share of egg deposition by survey component, before and after the 1985 dive-survey transition")

## records of more than one spawn event per location-year, before vs after
multi <- spawn_raw %>% filter(Region == SOG_REGION, StatisticalArea %in% OCC_AREAS) %>%
  count(LocationCode, Year) %>%
  group_by(Year) %>% summarise(pct_multi = 100*mean(n > 1), .groups = "drop") %>%
  filter(Year %in% 1951:2024) %>% arrange(Year)
sc_multi <- sctest(multi$pct_multi ~ 1, type = "supF")
bp_multi <- breakpoints(multi$pct_multi ~ 1, breaks = 1)
cat(sprintf("Multiple spawn records per location-year: structural break after %d (supF = %.0f, P %s)\n",
            multi$Year[bp_multi$breakpoints], sc_multi$statistic, fmt_p(sc_multi$p.value)))
cat(sprintf("  mean %% of location-years with >1 record: before %.0f%%, after %.0f%%\n",
            mean(multi$pct_multi[multi$Year <= multi$Year[bp_multi$breakpoints]]),
            mean(multi$pct_multi[multi$Year >  multi$Year[bp_multi$breakpoints]])))
```

---

## 2  Biomass does not explain occupancy once the phase is accounted for

```{r hysteresis-model}
m_hyst <- lm(occupancy ~ scale(SSB) + phase, data = d)
print(round(summary(m_hyst)$coefficients, 4))

p_ssb   <- summary(m_hyst)$coefficients["scale(SSB)", "Pr(>|t|)"]
p_phase <- summary(m_hyst)$coefficients[grep("^phase", rownames(summary(m_hyst)$coefficients)),
                                        "Pr(>|t|)"]
cat(sprintf("\nSSB: P %s | collapse phase: P %s\n", fmt_p(p_ssb), fmt_p(p_phase)))

cat(sprintf("Mean SSB: pre %.0f kt, post %.0f kt (post/pre = %.2f)\n",
            mean(pre$SSBk), mean(post$SSBk), mean(post$SSBk)/mean(pre$SSBk)))
cat(sprintf("Peak SSB: %.0f kt in %d; pre-collapse maximum %.0f kt in %d\n",
            max(d$SSBk), d$Year[which.max(d$SSBk)],
            max(pre$SSBk), pre$Year[which.max(pre$SSBk)]))
cat(sprintf("Minimum SSB: %.0f kt in %d, occupancy that year = %.0f%%\n",
            min(d$SSBk), d$Year[which.min(d$SSBk)], 100*d$occupancy[which.min(d$SSBk)]))
```

### Matched-biomass contrast

Because the two phases differ in mean biomass, the phase effect is re-examined over the
range of SSB actually observed before the collapse, so that pre- and post-collapse years are
contrasted at equivalent biomass.

```{r matched}
matched_contrast <- function(dat) {
  rng <- range(dat$SSB[dat$phase == "Pre-collapse"])
  a <- dat %>% filter(phase == "Pre-collapse")
  b <- dat %>% filter(phase == "Post-collapse", SSB >= rng[1], SSB <= rng[2])
  tibble(occ_pre = mean(a$occupancy), n_pre = nrow(a),
         occ_post_matched = mean(b$occupancy), n_post_matched = nrow(b))
}
mc <- matched_contrast(d)
cat(sprintf("At equivalent biomass: pre %.0f%% (n = %d) vs post %.0f%% (n = %d)\n",
            100*mc$occ_pre, mc$n_pre, 100*mc$occ_post_matched, mc$n_post_matched))
```

---

## 3  Supplementary Table S1 - sensitivity to rescaling the surface-survey index

Surface and dive surveys are not on the same scale: surface surveys used subjective
intensity categories before 1978 and systematically underestimate spawn width, a direct
multiplier on biomass. DFO estimates separate index scaling parameters for the two periods
(q₁, q₂). The comparison above is therefore repeated with the surface-era index (≤ 1987, DFO's q₁ period)
multiplied by a factor *k* spanning the full plausible range - from no correction (k = 1) up
to and beyond the value that **eliminates the apparent recovery entirely**, i.e. that makes
mean pre-collapse biomass equal mean post-collapse biomass.

If the decoupling were an artefact of the index scale, some value of *k* should make biomass
significant and close the occupancy gap. None does.

```{r tableS1}
k_equalise <- mean(post$SSB) / mean(pre$SSB)
cat(sprintf("k that equalises pre- and post-collapse mean SSB: %.2f\n", k_equalise))

k_grid <- sort(unique(round(c(1, 1.25, 1.5, k_equalise, 2, 2.5, 3), 2)))

tabS1 <- map_dfr(k_grid, function(k) {
  dk <- d %>% mutate(SSB = if_else(Year <= 1987, SSB * k, SSB), SSBk = SSB/1000)
  m  <- lm(occupancy ~ scale(SSB) + phase, data = dk)
  cf <- summary(m)$coefficients
  mc <- matched_contrast(dk)
  pk <- dk %>% filter(phase == "Pre-collapse"); qk <- dk %>% filter(phase == "Post-collapse")
  tibble(`Surface-era scaling k`     = k,
         `Mean pre-collapse SSB (kt)`  = round(mean(pk$SSBk)),
         `Mean post-collapse SSB (kt)` = round(mean(qk$SSBk)),
         `Post/pre biomass ratio`      = round(mean(qk$SSBk)/mean(pk$SSBk), 2),
         `SSB P`   = signif(cf["scale(SSB)","Pr(>|t|)"], 2),
         `Phase P` = signif(cf[grep("^phase", rownames(cf)),"Pr(>|t|)"], 2),
         `Occupancy pre (%)`               = round(100*mc$occ_pre),
         `Occupancy post, matched SSB (%)` = round(100*mc$occ_post_matched),
         `n matched post-collapse years`   = mc$n_post_matched)
})

save_tab(tabS1, "TableS1_survey_rescaling_sensitivity")

## Supplementary Table S7 (break and decoupling columns): the same statistics under three
## rules for which spawn records count as presence (series written by 01). 05 adds the GAMM.
occ_rules <- read_csv(file.path(DERIVED, "occupancy_rule_sensitivity.csv"), show_col_types = FALSE) %>%
  filter(Year %in% d$Year) %>%
  left_join(d %>% select(Year, SSB), by = "Year") %>%
  mutate(phase = factor(if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"),
                        levels = c("Pre-collapse", "Post-collapse")))
tabS7a <- occ_rules %>% group_by(rule) %>% group_modify(function(x, k) {
  x  <- arrange(x, Year)
  sc <- sctest(x$occupancy ~ 1, type = "supF")
  bp <- breakpoints(x$occupancy ~ 1, breaks = 1)
  ci <- tryCatch(confint(bp, breaks = 1)$confint, error = function(e) c(NA, NA, NA))
  cf <- summary(lm(occupancy ~ scale(SSB) + phase, data = x))$coefficients
  mc <- matched_contrast(x)
  tibble(Sections = x$n_sections[1],
         supF     = round(sc$statistic),
         `Break (95% CI)` = sprintf("after %d (%d-%d)", x$Year[bp$breakpoints], x$Year[ci[1]], x$Year[ci[3]]),
         `Occupancy pre (%)`  = round(100*mean(x$occupancy[x$phase == "Pre-collapse"])),
         `Occupancy post (%)` = round(100*mean(x$occupancy[x$phase == "Post-collapse"])),
         `Occupancy post, matched SSB (%)` = round(100*mc$occ_post_matched),
         `SSB P`   = signif(cf["scale(SSB)", "Pr(>|t|)"], 2),
         `Phase P` = signif(cf["phasePost-collapse", "Pr(>|t|)"], 2))
}) %>% ungroup()
save_tab(tabS7a, "TableS7_occupancy_rule_part_decoupling")
kable(tabS7a, caption = "Occupancy decoupling under three presence rules")
kable(tabS1, caption = paste(
  "Supplementary Table S1. Sensitivity of the biomass-occupancy decoupling to rescaling of",
  "the surface-survey era (<= 1987) spawn index. k = 1 is the published index;",
  sprintf("k = %.2f equalises mean pre- and post-collapse biomass, eliminating the", k_equalise),
  "apparent recovery. Occupancy is compared over the range of SSB observed before the collapse."))

## report what the table shows, rather than asserting non-significance
k_sig <- tabS1$`Surface-era scaling k`[tabS1$`SSB P` < 0.05]
cat(sprintf("\nAcross the full range of k: SSB P = %.3f-%.2f; %s; phase P %s throughout.\n",
            min(tabS1$`SSB P`), max(tabS1$`SSB P`),
            if (length(k_sig)) paste0("P < 0.05 only at k = ", paste(k_sig, collapse = ", "),
                                      " (within-phase SSB effect negative)") else "never significant",
            fmt_p(max(tabS1$`Phase P`))))
cat(sprintf("Occupancy at matched biomass: pre %d%%, post %d-%d%%.\n",
            unique(tabS1$`Occupancy pre (%)`)[1],
            min(tabS1$`Occupancy post, matched SSB (%)`),
            max(tabS1$`Occupancy post, matched SSB (%)`)))
```

---

## 4  Latitudinal extent: contraction, not a directional shift

```{r latitude}
occ_lat <- site_year %>%
  filter(Year %in% 1951:2025) %>%   # every recorded spawn, with or without a biomass estimate
  mutate(phase = if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"))

occ_lat %>% group_by(phase) %>%
  summarise(lat_min = round(min(Latitude), 2),
            q05     = round(quantile(Latitude, 0.05), 2),
            median  = round(median(Latitude), 2),
            q95     = round(quantile(Latitude, 0.95), 2),
            lat_max = round(max(Latitude), 2),
            .groups = "drop") %>%
  kable(caption = "Latitudinal distribution of occupied spawning locations, before and after the collapse")

## biomass-weighted centre of gravity - tests for a directional shift
cog <- site_year %>% filter(spawn_bio > 0) %>% group_by(Year) %>%
  summarise(cog_lat = weighted.mean(Latitude, spawn_bio), .groups = "drop") %>%
  mutate(phase = if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"))
tt <- t.test(cog_lat ~ phase, data = cog)
cat(sprintf("Biomass-weighted centre of gravity: pre %.2f degN, post %.2f degN, difference %.2f deg (t-test P %s)\n",
            mean(cog$cog_lat[cog$phase == "Pre-collapse"]),
            mean(cog$cog_lat[cog$phase == "Post-collapse"]),
            diff(rev(tapply(cog$cog_lat, cog$phase, mean))), fmt_p(tt$p.value)))
```

---

## 5  Temporal contraction of the spawning season (Supplementary Fig. S1)

The main text reports that spatial contraction was accompanied by temporal contraction, the
season "compressing from roughly four months to predominantly a single month". The statistic
behind that statement is computed here from the spawn-record start dates.

```{r season}
season <- spawn_raw %>%
  filter(Region == SOG_REGION, StatisticalArea %in% OCC_AREAS, !is.na(StartDate)) %>%
  mutate(start = suppressWarnings(as.Date(StartDate)),
         month = as.integer(format(start, "%m")),
         bio   = rowSums(cbind(Surface, Understory, Macrocystis), na.rm = TRUE)) %>%
  filter(!is.na(month), bio > 0, Year %in% 1951:2025) %>%
  mutate(phase = if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"))

season %>% group_by(phase, month) %>%
  summarise(biomass = sum(bio), .groups = "drop") %>%
  group_by(phase) %>% mutate(share = 100*biomass/sum(biomass)) %>%
  select(phase, month, share) %>%
  pivot_wider(names_from = phase, values_from = share) %>%
  arrange(month) %>% mutate(across(where(is.numeric), ~round(., 1))) %>%
  kable(caption = "Percentage of annual spawn biomass by calendar month, before and after the collapse")

## number of months needed to account for 90% of spawn biomass
n_months_90 <- season %>% group_by(phase, month) %>%
  summarise(b = sum(bio), .groups = "drop") %>%
  group_by(phase) %>% arrange(desc(b), .by_group = TRUE) %>%
  mutate(cum = cumsum(b)/sum(b)) %>%
  summarise(n_months = sum(lag(cum, default = 0) < 0.90), .groups = "drop")
print(n_months_90)

peak <- season %>% group_by(phase, month) %>% summarise(b = sum(bio), .groups = "drop") %>%
  group_by(phase) %>% mutate(share = b/sum(b)) %>% slice_max(share, n = 1)
cat(sprintf("\nPeak month share: pre-collapse month %d = %.0f%%; post-collapse month %d = %.0f%%\n",
            peak$month[peak$phase == "Pre-collapse"], 100*peak$share[peak$phase == "Pre-collapse"],
            peak$month[peak$phase == "Post-collapse"], 100*peak$share[peak$phase == "Post-collapse"]))
```

### How the season narrowed

The season was already March-centred before the collapse, so the change is not a shift of
the peak but the loss of the February and April shoulders. These are the three quantities
quoted in the Results.

```{r season-narrowing}
mshare <- season %>% group_by(phase, month) %>%
  summarise(bio = sum(bio), .groups = "drop") %>%
  group_by(phase) %>% mutate(share = 100*bio/sum(bio)) %>% ungroup()

for (ph in c("Pre-collapse","Post-collapse")) {
  s <- mshare %>% filter(phase == ph)
  cat(sprintf("%-14s months with >= 5%% of spawn biomass: %s | outside March: %.0f%%\n",
              ph, paste(sort(s$month[s$share >= 5]), collapse = ", "),
              sum(s$share[s$month != 3])))
}

## outer envelope: first to last spawn of the year, averaged within era
env <- season %>% mutate(doy = as.integer(format(start, "%j"))) %>%
  group_by(phase) %>%
  summarise(envelope_d = max(doy) - min(doy),
            spread_10_90 = quantile(doy, .9) - quantile(doy, .1), .groups = "drop")
print(env)
```

### Spawning waves within a site-year

A site-year can carry several spawn records. Most are concurrent - 62% of consecutive
records at the same site-year overlap or share a start date - and those are one spawning
episode mapped in pieces, which is what the per-site annual maximum removes. Records
separated by a week or more are a different quantity: they are successive spawning events at
the same place. Hourston & Haegele (1980) describe exactly this, "two waves of spawning" at
a site, alongside the individual spawning act.

The count of such waves is confounded two ways, and both are controlled below. Survey method
changed in 1988, and the season itself shortened, which mechanically leaves less room for a
second wave. `WAVE_GAP` is the separation in days required to call two records distinct.

```{r spawn-waves}
WAVE_GAP <- 7

wave_events <- spawn_raw %>%
  filter(Region == SOG_REGION, StatisticalArea %in% OCC_AREAS, !is.na(StartDate), Year %in% 1951:2025) %>%
  mutate(start = suppressWarnings(as.Date(StartDate)),
         end   = suppressWarnings(as.Date(EndDate)),
         bio   = rowSums(cbind(Surface, Understory, Macrocystis), na.rm = TRUE)) %>%
  filter(!is.na(start)) %>%   # spawn events, including records without a biomass estimate
  mutate(end = if_else(is.na(end), start, end)) %>%
  arrange(LocationCode, Year, start)

count_waves <- function(gap) {
  wave_events %>% group_by(LocationCode, Year) %>%
    summarise(Area   = first(StatisticalArea),
              Method = names(sort(table(Method), decreasing = TRUE))[1],
              n_rec  = n(),
              n_waves = 1L + sum(as.numeric(start[-1] - head(end, -1)) >= gap),
              .groups = "drop") %>%
    mutate(multi = as.integer(n_waves > 1),
           post  = as.integer(Year >= BREAK_YEAR))
}

## annual season breadth, computed across all sites so it is independent of any
## single site-year's wave count
breadth <- wave_events %>% mutate(doy = as.integer(format(start, "%j"))) %>%
  group_by(Year) %>%
  summarise(season_breadth = quantile(doy, .9) - quantile(doy, .1), .groups = "drop")

wv <- count_waves(WAVE_GAP) %>% left_join(breadth, by = "Year")

cat("Threshold sensitivity, surface surveys only (P that a site-year shows > 1 wave):\n")
for (g in c(5, 7, 10, 14)) {
  s <- count_waves(g) %>% filter(Method == "Surface")
  cat(sprintf("  >= %2d d apart : pre %4.1f%% -> post %4.1f%%\n", g,
              100*mean(s$multi[s$post == 0]), 100*mean(s$multi[s$post == 1])))
}

cat(sprintf("\nSeason breadth (10-90%% spread of spawn start): pre %.0f d -> post %.0f d\n",
            mean(breadth$season_breadth[breadth$Year <  BREAK_YEAR]),
            mean(breadth$season_breadth[breadth$Year >= BREAK_YEAR])))

## Does the era effect survive the two confounds?
## n_rec is deliberately NOT a covariate: n_waves is derived from those same
## records, so adjusting for it conditions on a mediator and inflates the effect.
cat("\nLogistic models for P(> 1 wave), all methods:\n")
for (f in c("multi ~ post",
            "multi ~ post + season_breadth",
            "multi ~ post + season_breadth + factor(Method)")) {
  m  <- glm(as.formula(f), data = wv, family = binomial)
  cf <- summary(m)$coefficients["post", ]
  cat(sprintf("  %-46s OR = %.2f, P %s\n", f, exp(cf[1]), fmt_p(cf[4])))
}

m_wave <- glm(multi ~ post + season_breadth + factor(Method), data = wv, family = binomial)
cf <- summary(m_wave)$coefficients["post", ]

## Quote the descriptive percentages from the surface-only subset, where method is
## held constant by construction, and the odds ratio from the adjusted model on the
## full data. Mixing the two - surface percentages beside an all-methods OR - would
## describe two different subsets in one sentence.
sf <- wv %>% filter(Method == "Surface")
cat(sprintf("\nDescriptive (surface only, method constant): pre %.0f%% (n = %d) -> post %.0f%% (n = %d)\n",
            100*mean(sf$multi[sf$post == 0]), sum(sf$post == 0),
            100*mean(sf$multi[sf$post == 1]), sum(sf$post == 1)))
cat(sprintf("Descriptive (all methods, unadjusted):        pre %.0f%% -> post %.0f%%\n",
            100*mean(wv$multi[wv$post == 0]), 100*mean(wv$multi[wv$post == 1])))
cat(sprintf("Adjusted for season breadth + method (all data): OR = %.2f, P %s\n",
            exp(cf[1]), fmt_p(cf[4])))

save_tab(wv %>% count(post, multi) %>% mutate(gap_days = WAVE_GAP), "spawning_waves_by_era")
```

**What this supports and what it does not.** The decline is robust to the separation
threshold and survives adjustment for both season breadth and survey method, so it is not
merely the season shortening and not merely the dive-survey transition. It cannot, however,
distinguish the same fish returning from a second group arriving: spawn surveys record eggs,
not individuals. It is a loss of within-season spawning structure at a site, reported
alongside the temporal contraction rather than as an independent mechanism. Post-collapse
surface-survey site-years are also fewer and are the subset that did not warrant a dive
survey, which is why the method-adjusted model is the one quoted.

```{r figS1, fig.width=9, fig.height=6}
## Supplementary Figure S1 - each point is a spawn event, placed by its start day
## of year and scaled by biomass, so the compression of the season is visible
## directly rather than only through the monthly shares tabulated above.
season_pts <- spawn_raw %>%
  filter(Region == SOG_REGION, StatisticalArea %in% OCC_AREAS, !is.na(StartDate)) %>%
  mutate(start  = suppressWarnings(as.Date(StartDate)),
         doy    = as.integer(format(start, "%j")),
         bio    = rowSums(cbind(Surface, Understory, Macrocystis), na.rm = TRUE),
         area_f = factor(as.character(StatisticalArea), levels = names(area_cols))) %>%
  filter(!is.na(doy), Year %in% 1951:2025, !is.na(area_f))

pS1 <- ggplot(season_pts, aes(Year, doy)) +
  geom_rect(data = shade_df, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "grey85", alpha = 0.45) +
  band_closure() +
  geom_point(aes(size = bio, colour = area_f), alpha = 0.65) +
  scale_size_continuous(name = "Spawning biomass (t)", range = c(0.6, 5.5),
                        labels = label_number(big.mark = ",")) +
  scale_colour_manual(values = area_cols, name = "Management area") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  labs(x = "Year", y = "Day of year") +
  common_panel_legend(pos = "bottom", just = "center") +
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1), nrow = 1),
         size   = "none")

save_fig(pS1, "FigureS1_temporal_contraction")
pS1
```

---

## 6  Figure 1 - the assembled panels

All six panels of Figure 1 are built here from the derived series, so the figure and the
statistics above can never diverge. Panels **a** and **b** plot the same `SSB` and
`occupancy` columns used by every model in this chapter: SSB is the sum of all spawn records
(the DFO spawn index; §1 of chapter 01) and occupancy is over the `r N_all` historically-used
sections, counting every recorded spawn, including records without a biomass estimate.

```{r fig1-data}
## Hysteresis needs the long era labels that carry the year ranges into the legend.
d_hyst <- d %>%
  mutate(phase2 = factor(if_else(Year < BREAK_YEAR, PHASE_LEVELS[1], PHASE_LEVELS[2]),
                         levels = PHASE_LEVELS))
pre_h  <- d_hyst %>% filter(phase2 == PHASE_LEVELS[1])
post_h <- d_hyst %>% filter(phase2 == PHASE_LEVELS[2])
pre_hull  <- pre_h [chull(pre_h$SSBk,  pre_h$occupancy),  ]
post_hull <- post_h[chull(post_h$SSBk, post_h$occupancy), ]

## Latitudinal extent: occupied locations, with 5th/95th percentile edges smoothed.
lat_pts <- site_year %>%
  filter(Year %in% 1951:2025) %>%
  mutate(area_f = factor(as.character(StatisticalArea), levels = names(area_cols)))

lat_q <- lat_pts %>% group_by(Year) %>%
  summarise(lat05 = quantile(Latitude, 0.05, na.rm = TRUE),
            lat95 = quantile(Latitude, 0.95, na.rm = TRUE), .groups = "drop")

g05 <- gam(lat05 ~ s(Year, k = 16), data = lat_q)
g95 <- gam(lat95 ~ s(Year, k = 10), data = lat_q)
lat_grid <- tibble(Year = seq(min(lat_q$Year), max(lat_q$Year), length.out = 200))
.f05 <- predict(g05, lat_grid, se.fit = TRUE)
.f95 <- predict(g95, lat_grid, se.fit = TRUE)
lat_edge <- bind_rows(
  tibble(Year = lat_grid$Year, fit = .f05$fit, se = .f05$se.fit, edge = "q05"),
  tibble(Year = lat_grid$Year, fit = .f95$fit, se = .f95$se.fit, edge = "q95"))

## Catch by management area, with the TAC overlaid.
catch_area <- read_csv(file.path(DERIVED, "catch_decomposition.csv"), show_col_types = FALSE) %>%
  group_by(Year, area) %>%
  summarise(catch_kt = sum(catch_t, na.rm = TRUE)/1000, .groups = "drop") %>%
  mutate(area_f = factor(as.character(area), levels = names(area_cols))) %>%
  filter(!is.na(area_f), Year %in% 1951:2025)

tac <- read_csv(F_TAC, show_col_types = FALSE) %>%
  rename_with(trimws) %>%
  transmute(Year = as.numeric(New_Year), TAC_kt = as.numeric(TAC_metric_tons)/1000) %>%
  filter(!is.na(Year), Year %in% 1951:2025)

## Landed value: nominal and deflated to constant 2024 CAD.
value_two <- read_excel(F_VALUE, sheet = 1) %>%
  rename_with(trimws) %>%
  transmute(Year     = suppressWarnings(as.numeric(Year)),
            Nominal  = suppressWarnings(as.numeric(value_per_ton)),
            Real2024 = suppressWarnings(as.numeric(value_per_ton_real2024))) %>%
  filter(!is.na(Year), Year %in% 1950:2025) %>%
  group_by(Year) %>% summarise(across(everything(), ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  mutate(across(c(Nominal, Real2024), ~ifelse(is.nan(.x) | .x <= 0, NA_real_, .x)))

cat(sprintf("Fig 1 inputs | SSB %d-%d | catch areas %d | TAC %d yr | value %d yr\n",
            min(d$Year), max(d$Year), n_distinct(catch_area$area_f),
            nrow(tac), sum(!is.na(value_two$Real2024))))
```

```{r fig1-panels, fig.width=16, fig.height=15}
shade_layer <- geom_rect(data = shade_df,
                         aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
                         inherit.aes = FALSE, fill = "grey85", alpha = 0.45)

## ---- a  Spawning stock biomass ---------------------------------------------
p1a <- ggplot(d, aes(Year, SSBk)) +
  shade_layer +
  geom_line(linewidth = 0.9, lineend = "round") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10), se = TRUE,
              colour = "darkblue", fill = "grey75", linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  labs(x = "Year", y = "SSB (×1000 tonnes)") +
  common_panel

## ---- b  Spawning occupancy --------------------------------------------------
p1b <- ggplot(d, aes(Year, occupancy)) +
  shade_layer +
  geom_line(linewidth = 0.9, lineend = "round") +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 10), se = TRUE,
              colour = "darkblue", fill = "grey75", linewidth = 1) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15)),
                     labels = percent_format(accuracy = 1)) +
  labs(x = "Year", y = "Spawning occupancy") +
  common_panel

## ---- c  Biomass-occupancy hysteresis ---------------------------------------
p1c <- ggplot(d_hyst, aes(SSBk, occupancy)) +
  geom_polygon(data = pre_hull,  aes(SSBk, occupancy), inherit.aes = FALSE,
               fill = phase_hull[[1]], alpha = 0.28) +
  geom_polygon(data = post_hull, aes(SSBk, occupancy), inherit.aes = FALSE,
               fill = phase_hull[[2]], alpha = 0.22) +
  geom_path(colour = "grey60", linewidth = 0.4, alpha = 0.55) +
  geom_point(aes(fill = phase2, colour = phase2), shape = 21, size = 3.4, stroke = 0.5) +
  scale_fill_manual(values = phase_fill, name = NULL) +
  scale_colour_manual(values = phase_edge, name = NULL, guide = "none") +
  scale_x_continuous(labels = label_number(big.mark = ","),
                     expand = expansion(mult = c(0.02, 0.04))) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(x = "SSB (×1000 tonnes)", y = "Spawning occupancy") +
  theme_science +
  theme(legend.position = c(0.98, 0.98), legend.justification = c(1, 1),
        legend.background = element_rect(fill = alpha("white", 0.6), colour = NA),
        legend.key = element_blank(), panel.grid = element_blank())

## ---- d  Latitudinal extent --------------------------------------------------
p1d <- ggplot(lat_pts, aes(Year, Latitude)) +
  shade_layer +
  geom_point(aes(size = spawn_bio, colour = area_f), alpha = 0.65) +
  geom_ribbon(data = lat_edge, aes(x = Year, ymin = fit - 1.96*se, ymax = fit + 1.96*se,
                                   group = edge),
              inherit.aes = FALSE, fill = "grey30", alpha = 0.18) +
  geom_line(data = lat_edge, aes(x = Year, y = fit, group = edge),
            inherit.aes = FALSE, colour = "grey25", linewidth = 0.9) +
  scale_size_continuous(name = "Spawning biomass (t)", range = c(0.6, 5.5),
                        labels = label_number(big.mark = ",")) +
  scale_colour_manual(values = area_cols, name = "Management area", drop = TRUE) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  labs(x = "Year", y = "Latitude (°N)") +
  ## the point cloud fills the panel, so this legend goes outside it
  common_panel_legend(pos = "bottom", just = "center") +
  guides(colour = guide_legend(override.aes = list(size = 5, alpha = 1), nrow = 2),
         size   = "none")

## ---- e  Catch by management area, with TAC ---------------------------------
p1e <- ggplot(catch_area, aes(Year, catch_kt, fill = area_f)) +
  shade_layer + band_closure() +
  geom_area(colour = NA) +
  geom_line(data = tac, aes(Year, TAC_kt), inherit.aes = FALSE,
            colour = "#C0392B", linewidth = 1) +
  scale_fill_manual(values = area_cols, name = "Management area") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Year", y = "Catch (×1000 tonnes)") +
  common_panel_legend(pos = c(0.99, 0.99), just = c(1, 1)) +
  guides(fill = guide_legend(nrow = 3))

## ---- f  Landed value per tonne ---------------------------------------------
p1f <- ggplot(value_two, aes(Year)) +
  shade_layer + band_closure() +
  geom_line(aes(y = Nominal/1000),  colour = "grey55", linewidth = 0.9, linetype = "dashed") +
  geom_line(aes(y = Real2024/1000), colour = "black",  linewidth = 0.9) +
  annotate("text", x = 1976.5, y = max(value_two$Real2024, na.rm = TRUE)/1000,
           label = "Herring\nGold Rush", vjust = 0.9, hjust = 1, size = 5.5, lineheight = 0.95) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.22))) +
  labs(x = "Year", y = "Value per tonne (×1000 CAD)") +
  common_panel

for (nm in c("a","b","c","d","e","f")) {
  save_fig(get(paste0("p1", nm)), paste0("Fig1", nm, "_",
    c(a="SSB", b="occupancy", c="hysteresis", d="latitudinal_extent",
      e="catch_by_area", f="value_per_tonne")[[nm]]))
}

fig1 <- (p1a | p1d) / (p1b | p1e) / (p1c | p1f) +
  plot_annotation(tag_levels = list(c("a","d","b","e","c","f"))) &
  theme(plot.tag = element_text(size = 26, face = "bold"))

save_fig(fig1, "Figure1_biomass_occupancy_decoupling", width = 18, height = 16)
fig1
```

---

## Session information

```{r session}
sessionInfo()
```


---

## Sensitivity: SSB as per-location annual maxima

Main-text SSB is the sum of all spawn records (DFO spawn index). Because multiple records per
location-year were far more common before the collapse, the biomass-occupancy model is repeated
with SSB_max, the sum of per-location annual maxima.

```{r ssb-max-sensitivity}
ann_s <- read_csv(file.path(DERIVED, "annual_series.csv"), show_col_types = FALSE) %>%
  mutate(phase = factor(if_else(Year < BREAK_YEAR, "Pre-collapse", "Post-collapse"),
                        levels = c("Pre-collapse", "Post-collapse")))
m_max <- lm(occupancy ~ scale(SSB_max) + phase, data = ann_s)
rng_max <- range(ann_s$SSB_max[ann_s$phase == "Pre-collapse"])
cat(sprintf("SSB_max sensitivity: SSB P = %.2f | phase P %s | occupancy pre %.0f%% vs post at matched SSB_max %.0f%%\n",
            summary(m_max)$coefficients["scale(SSB_max)", "Pr(>|t|)"],
            fmt_p(summary(m_max)$coefficients["phasePost-collapse", "Pr(>|t|)"]),
            100 * mean(ann_s$occupancy[ann_s$phase == "Pre-collapse"]),
            100 * mean(ann_s$occupancy[ann_s$phase == "Post-collapse" &
                                         ann_s$SSB_max >= rng_max[1] & ann_s$SSB_max <= rng_max[2]])))
loc_multi <- read_csv(file.path(DERIVED, "site_year_spawn_all.csv"), show_col_types = FALSE) %>%
  mutate(era = if_else(Year < BREAK_YEAR, "pre", "post")) %>% group_by(era) %>%
  summarise(pct_multi_record = round(100 * mean(n_records > 1)), .groups = "drop")
print(loc_multi)
```
