---
title: "04 — Size-selective fishing and the restructuring of population demography"
subtitle: "Results 2.2: gear selectivity, length-at-age, age-structure change points (Supplementary Fig. S3)"
author: "Sarah Ohayon, Jacob T. Dingwall, Amanda E. Bates"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")
source("00_setup.R")
library(mgcv); library(segmented); library(patchwork); library(knitr); library(scales)
```

```{r load}
bio_src   <- read_csv(file.path(DERIVED, "biosample_annual_by_source.csv"), show_col_types = FALSE)
test_age  <- read_csv(file.path(DERIVED, "test_fishery_age.csv"),           show_col_types = FALSE)
len_age   <- read_csv(file.path(DERIVED, "length_at_age.csv"),              show_col_types = FALSE)
catchdec  <- read_csv(file.path(DERIVED, "catch_decomposition.csv"),        show_col_types = FALSE)
```

---

## 1  The fishery transition: season, gear and demographic target

```{r fishery-transition}
catchdec %>%
  filter(area %in% BIO_AREAS, fishery %in% c("Reduction","Roe","Food and bait")) %>%
  group_by(fishery, gear_grp) %>%
  summarise(catch_kt = round(sum(catch_t, na.rm = TRUE)/1000), .groups = "drop") %>%
  pivot_wider(names_from = gear_grp, values_from = catch_kt, values_fill = 0) %>%
  kable(caption = "Total catch (kt) by fishery type and gear, statistical areas 13-18")

roe <- catchdec %>% filter(area %in% BIO_AREAS, fishery == "Roe",
                           gear_grp %in% c("Seine","Gillnet"), Year %in% 1972:2025)
gn_share_overall <- with(roe, sum(catch_t[gear_grp == "Gillnet"], na.rm = TRUE) /
                              sum(catch_t, na.rm = TRUE))
gn_annual <- roe %>% group_by(Year, gear_grp) %>%
  summarise(c = sum(catch_t, na.rm = TRUE), .groups = "drop") %>%
  group_by(Year) %>% summarise(share = c[gear_grp == "Gillnet"]/sum(c), .groups = "drop")

cat(sprintf("Gillnets took %.0f%% of the roe catch 1972-2025; annual share IQR %.0f-%.0f%%\n",
            100*gn_share_overall,
            100*quantile(gn_annual$share, .25, na.rm = TRUE),
            100*quantile(gn_annual$share, .75, na.rm = TRUE)))
```

---

## 2  Age composition differs among sampling sources (Fig. 2b)

The question is whether the sources are sampling the same age structure or different ones.
Each row of the data is one source-year, with counts of old (age ≥ 5) and young fish.

**Why not a plain binomial model.** A binomial model assumes every fish is an independent draw,
so its variance is fixed by the counts. Fish are caught in schools and sampled in sets, so the
real year-to-year scatter is far larger: residual deviance/df is ~100, i.e. the data are ~100×
overdispersed. Under that assumption the standard errors are ~10× too small and AIC differences
become meaningless (the earlier dAIC = 65,403). Two corrections are used instead:

1. **Quasibinomial GAM** — the same model, but the dispersion is estimated from the data rather
   than fixed at 1, and the common-trend vs source-specific-trend comparison is an F test.
2. **Binomial GAM with an observation-level random effect** (`s(obs, bs = "re")`) — each
   source-year gets its own random deviation, which absorbs the extra scatter. With one shared
   temporal smooth, the source coefficients are the difference in *level* between sources,
   reported as odds ratios relative to the test fishery (the closest index of the spawning
   population). AIC is not used with this model: the observation-level effect absorbs source
   differences, so conditional AIC is unreliable for comparing fixed effects; Wald tests are used.

Caveat: reduction samples (1946-1972) do not overlap in time with the test fishery (1975-), so
that contrast relies on the shared temporal smooth.

```{r source-gam}
gd <- bio_src %>%
  mutate(src = factor(src, levels = c("Reduction","Roe seine","Roe gillnet",
                                      "Food and bait","Test fishery"))) %>%
  filter(!is.na(src)) %>%
  mutate(obs = factor(row_number()))

## plain binomial, kept only to document the overdispersion
m_source_bin <- gam(cbind(n_old, n_young) ~ src + s(Year, by = src, k = 8),
                    family = binomial, data = gd, method = "REML")
cat(sprintf("Source-years: %d | fish: %s | binomial residual deviance/df = %.1f (1 = no overdispersion)\n",
            nrow(gd), format(sum(gd$n_fish), big.mark = ","),
            deviance(m_source_bin) / df.residual(m_source_bin)))

## 1  quasibinomial: common trend vs source-specific trends
## trends are regression splines with a fixed 7 df (k = 8, fx = TRUE), the model drawn in Fig. 2b:
## with automatic smoothing the ~100x overdispersion flattens every source trend
m_common <- gam(cbind(n_old, n_young) ~ s(Year, k = 8, fx = TRUE),
                family = quasibinomial, data = gd)
m_source <- gam(cbind(n_old, n_young) ~ src + s(Year, by = src, k = 8, fx = TRUE),
                family = quasibinomial, data = gd)
f_src <- anova(m_common, m_source, test = "F")
cat(sprintf("Quasibinomial: dispersion = %.1f | source-specific vs common trend F = %.1f, P %s\n",
            summary(m_source)$scale, f_src$F[2], fmt_p(f_src$`Pr(>F)`[2])))
cat(sprintf("Deviance explained: common %.1f%%, source-specific %.1f%%\n",
            100*summary(m_common)$dev.expl, 100*summary(m_source)$dev.expl))

## 2  shared trend + source levels + observation-level random effect: odds ratios vs test fishery
gd_or <- gd %>% mutate(src = relevel(src, ref = "Test fishery"))
m_level <- gam(cbind(n_old, n_young) ~ src + s(Year, k = 8) + s(obs, bs = "re"),
               family = binomial, data = gd_or, method = "REML")
pt <- summary(m_level)$p.table[-1, , drop = FALSE]
src_or <- tibble(Source = sub("^src", "", rownames(pt)),
                 `OR vs test fishery` = exp(pt[, 1]),
                 lower = exp(pt[, 1] - 1.96 * pt[, 2]),
                 upper = exp(pt[, 1] + 1.96 * pt[, 2]),
                 P = pt[, 4])
save_tab(src_or, "prop_old_source_odds_ratios")
kable(src_or, digits = c(0, 2, 2, 2, 4),
      caption = "Odds of an old fish (age >= 5) relative to the test fishery (shared trend, observation-level random effect)")
```

```{r source-summaries}
src_summary <- gd %>% group_by(src) %>%
  summarise(years   = paste(range(Year), collapse = "-"),
            n_years = n(),
            prop_old_count  = sprintf("%.2f +/- %.2f", mean(prop_old_n), sd(prop_old_n)),
            prop_old_weight = sprintf("%.2f +/- %.2f", mean(prop_old_wt, na.rm = TRUE),
                                      sd(prop_old_wt, na.rm = TRUE)),
            .groups = "drop")
save_tab(src_summary, "prop_old_by_sampling_source")
kable(src_summary,
      caption = "Proportion of old fish (age >= 5) by sampling source, mean +/- SD across years")

## the roe-gillnet trajectory quoted in the Results
rg <- gd %>% filter(src == "Roe gillnet") %>% arrange(Year)
cat(sprintf("Roe gillnet: %.2f in %d rising to %.2f by %d; mean over 1980-2024 = %.2f +/- %.2f\n",
            rg$prop_old_n[1], rg$Year[1],
            max(rg$prop_old_n[rg$Year <= 1980]), rg$Year[which.max(replace(rg$prop_old_n, rg$Year > 1980, NA))],
            mean(rg$prop_old_n[rg$Year >= 1980]), sd(rg$prop_old_n[rg$Year >= 1980])))

## fold-change in selectivity, by weight
red_wt <- gd %>% filter(src == "Reduction")   %>% pull(prop_old_wt)
gil_wt <- gd %>% filter(src == "Roe gillnet") %>% pull(prop_old_wt)
cat(sprintf("By weight: reduction %.2f +/- %.2f -> roe gillnet %.2f +/- %.2f (%.1f-fold)\n",
            mean(red_wt, na.rm = TRUE), sd(red_wt, na.rm = TRUE),
            mean(gil_wt, na.rm = TRUE), sd(gil_wt, na.rm = TRUE),
            mean(gil_wt, na.rm = TRUE)/mean(red_wt, na.rm = TRUE)))
```

---

## 3  Length-at-age (Fig. 2c)

```{r length-gam}
la <- len_age %>% mutate(age_f = factor(age))
m_len_common <- gam(mean_len_mm ~ age_f + s(Year, k = 8), weights = n,
                    data = la, method = "REML")
m_len_byage  <- gam(mean_len_mm ~ age_f + s(Year, by = age_f, k = 8), weights = n,
                    data = la, method = "REML")
cat(sprintf("Length-at-age: common smooth AIC = %.0f, age-specific smooths AIC = %.0f (dAIC = %.0f)\n",
            AIC(m_len_common), AIC(m_len_byage), AIC(m_len_common) - AIC(m_len_byage)))

## change in mean length-at-age between eras, per age class
la %>% mutate(era = case_when(Year <= 1979 ~ "<=1979",
                              Year <= 1983 ~ "1980-1983",
                              TRUE         ~ ">=1984")) %>%
  group_by(age, era) %>%
  summarise(mean_cm = round(weighted.mean(mean_len_mm, n)/10, 1), .groups = "drop") %>%
  pivot_wider(names_from = era, values_from = mean_cm) %>%
  mutate(`change 1980-83 vs <=1979 (cm)` = `1980-1983` - `<=1979`,
         `change >=1984 vs <=1979 (cm)`  = `>=1984`   - `<=1979`) %>%
  kable(caption = "Mean length-at-age (cm) by age class and era, sample-size weighted")
```

---

## 4  Test-fishery age structure: the three-phase contrast

The test fishery began in 1975 and samples during the spawning season, and so is the closest
available fishery-independent index of the spring spawning population. Mean age and the
proportion of old fish are contrasted across three phases: pre-collapse (≤ 1983), the
collapse itself (1984–1986) and the post-collapse period (≥ 1987).

A one-way ANOVA is used rather than a change-point test because the trajectory is a drop
**and a partial recovery**, which a single-breakpoint model cannot represent; on mean age the
change-point test returns a false negative for exactly that reason (Section 5).

```{r three-phase}
ta <- test_age %>% mutate(phase = factor(PHASE3(Year),
                                         levels = c("Pre-collapse","Collapse","Post-collapse")))

phase_test <- function(var, label) {
  f  <- as.formula(paste(var, "~ phase"))
  a  <- aov(f, data = ta); s <- summary(a)[[1]]
  mu <- ta %>% group_by(phase) %>%
    summarise(m = mean(.data[[var]]), sd = sd(.data[[var]]), n = n(), .groups = "drop")
  cat(sprintf("\n== %s ==\n", label))
  for (i in seq_len(nrow(mu)))
    cat(sprintf("  %-14s %.2f +/- %.2f (n = %d years)\n",
                mu$phase[i], mu$m[i], mu$sd[i], mu$n[i]))
  cat(sprintf("  ANOVA: F(%d,%d) = %.2f, P %s\n",
              s$Df[1], s$Df[2], s$`F value`[1], fmt_p(s$`Pr(>F)`[1])))
  tibble(Metric = label,
         `Pre-collapse`  = sprintf("%.2f +/- %.2f", mu$m[1], mu$sd[1]),
         `Collapse (1984-1986)` = sprintf("%.2f +/- %.2f", mu$m[2], mu$sd[2]),
         `Post-collapse` = sprintf("%.2f +/- %.2f", mu$m[3], mu$sd[3]),
         df = sprintf("%d, %d", s$Df[1], s$Df[2]),
         F  = round(s$`F value`[1], 2),
         P  = signif(s$`Pr(>F)`[1], 3))
}

three_phase <- bind_rows(
  phase_test("mean_age",  "Mean age (years)"),
  phase_test("prop_age5", "Proportion age >= 5"),
  phase_test("prop_age4", "Proportion age >= 4"),
  phase_test("prop_juv",  "Proportion age 1-3"))
save_tab(three_phase, "test_fishery_three_phase_contrasts")
kable(three_phase, caption = "Test-fishery age structure across the three phases (one-way ANOVA)")

cat(sprintf("\nTest-fishery series: %d years, %d-%d\n",
            nrow(ta), min(ta$Year), max(ta$Year)))
```

---

## 5  Change-point analysis of age structure (Supplementary Fig. S3)

Piecewise regression with a single breakpoint estimated from a 1986 starting value; the
Davies test asks whether a change in slope is present at all. Proportions are fitted as
binomial models weighted by the number of aged fish.

```{r changepoints, fig.width=11, fig.height=13}
wt_age <- ta$n_fish

fit_cp <- function(model, label) {
  seg <- try(segmented(model, seg.Z = ~Year, psi = list(Year = 1986)), silent = TRUE)
  if (inherits(seg, "try-error")) return(NULL)
  ps <- summary(seg)$psi
  dv <- tryCatch(davies.test(model, ~Year), error = function(e) NULL)
  list(seg = seg,
       row = tibble(Metric = label,
                    `Break year` = round(ps[, "Est."], 1),
                    SE           = round(ps[, "St.Err"], 2),
                    `95% CI`     = sprintf("%.1f-%.1f",
                                           ps[,"Est."] - 1.96*ps[,"St.Err"],
                                           ps[,"Est."] + 1.96*ps[,"St.Err"]),
                    `Davies P`   = if (is.null(dv)) NA_real_ else signif(dv$p.value, 3)),
       fitted = if (inherits(model, "glm") && family(model)$family == "binomial")
                  fitted(seg, type = "response") else fitted(seg))
}

m_mean  <- lm(mean_age ~ Year, data = ta)
m_a6    <- glm(prop_age6 ~ Year, family = binomial, data = ta, weights = wt_age)
m_a5    <- glm(prop_age5 ~ Year, family = binomial, data = ta, weights = wt_age)
m_a4    <- glm(prop_age4 ~ Year, family = binomial, data = ta, weights = wt_age)
m_juv   <- glm(prop_juv  ~ Year, family = binomial, data = ta, weights = wt_age)

cps <- list(
  `Mean age`            = fit_cp(m_mean, "Mean age (years)"),
  `Proportion age >= 6` = fit_cp(m_a6,   "Proportion age >= 6"),
  `Proportion age >= 5` = fit_cp(m_a5,   "Proportion age >= 5"),
  `Proportion age >= 4` = fit_cp(m_a4,   "Proportion age >= 4"),
  `Proportion age 1-3`  = fit_cp(m_juv,  "Proportion age 1-3"))

cp_tab <- map_dfr(cps, "row")
save_tab(cp_tab, "age_structure_changepoints")
kable(cp_tab, caption = "Change-point analysis of test-fishery age structure (segmented regression, Davies test)")
```

```{r cp-figure, fig.width=10, fig.height=12}
yvars <- c("mean_age","prop_age6","prop_age5","prop_age4","prop_juv")
ylabs <- c("Mean age (years)","Proportion age ≥ 6","Proportion age ≥ 5",
           "Proportion age ≥ 4","Proportion age 1–3")

## Panel a (mean age) shows the three-phase contrast reported in the Results, not a change
## point: mean age falls and then partially recovers, so a single break describes it poorly
## and the Davies test does not reject. Panels b-e keep the change-point fits.
ph_means <- ta %>%
  mutate(ph = factor(PHASE3(Year), levels = c("Pre-collapse","Collapse","Post-collapse"))) %>%
  group_by(ph) %>%
  summarise(m = mean(mean_age), se = sd(mean_age)/sqrt(n()),
            x0 = min(Year), x1 = max(Year), .groups = "drop")

panels <- map(seq_along(cps), function(i) {
  cp <- cps[[i]]; v <- yvars[i]
  bp <- as.numeric(summary(cp$seg)$psi[, "Est."])
  dv <- cp$row$`Davies P`
  p <- ggplot(ta, aes(Year, .data[[v]])) +
    band_postcoll(max(ta$Year)) + band_closure() +
    geom_line(colour = "grey35", linewidth = 0.7) +
    geom_point(aes(size = n_fish), colour = "grey30", alpha = 0.7, shape = 16)
  if (i == 1) {
    p <- p +
      geom_rect(data = ph_means, inherit.aes = FALSE,
                aes(xmin = x0, xmax = x1, ymin = m - se, ymax = m + se),
                fill = col_old, alpha = 0.15) +
      geom_segment(data = ph_means, inherit.aes = FALSE,
                   aes(x = x0, xend = x1, y = m, yend = m),
                   colour = col_old, linewidth = 1.2) +
      geom_text(data = ph_means, inherit.aes = FALSE,
                aes(x = (x0 + x1)/2,
                    y = if_else(ph == "Collapse", m - se, m + se),
                    vjust = if_else(ph == "Collapse", 1.8, -0.8),
                    label = sprintf("%.2f", m)),
                size = 3.4, colour = col_old) +
      annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.5, size = 3.4,
               label = "three-phase means \u00b1 SE (pre <= 1983, collapse 1984-1986, post >= 1987)")
  } else {
    p <- p +
      geom_line(aes(y = cp$fitted), colour = col_old, linewidth = 1.1) +
      geom_vline(xintercept = bp, linetype = "dashed", colour = "grey25") +
      annotate("text", x = -Inf, y = Inf, hjust = -0.08, vjust = 1.5, size = 3.4,
               label = sprintf("break %.0f; Davies P %s", bp,
                               if (is.na(dv)) "n/a" else if (dv < 1e-4) "< 0.0001"
                               else sprintf("= %.4f", dv)))
  }
  p +
    scale_size_area(name = "Fish aged", max_size = 4, labels = label_number(big.mark = ",")) +
    labs(x = NULL, y = ylabs[i]) +
    theme_herring(11) +
    theme(legend.position = if (i == 1) c(0.99, 0.02) else "none",
          legend.justification = c(1, 0))
})

figS3 <- wrap_plots(panels, ncol = 1) +
  plot_annotation(tag_levels = "a",
    title = "Supplementary Figure S3. Age structure of the Strait of Georgia test fishery",
    theme = theme(plot.title = element_text(face = "bold", size = 12)))
save_fig(figS3, "FigureS3_age_structure_changepoints", width = 9, height = 14)
figS3
```

**Reading the panels.** The break in the proportion metrics falls at 1985–1986 under every
definition of old fish, and at the same date in the mirror-image rise of juveniles and
first-time spawners. Mean age is the one metric on which the change-point test is
uninformative: it drops and then partially recovers, so a single change in slope describes
it poorly and the Davies test does not reject. That is why panel a shows the three-phase
contrast reported in Section 4 (pre-collapse, the 1984-1986 collapse years and post-collapse)
rather than a change-point fit.

---

## 6  Figure 2 — the assembled panels

The four panels are built from the same objects the statistics above are computed from:
`catchdec` (a), `gd` (b), `la` (c) and `ta` (d). Every panel carries the same two bands —
red for the 1968–1971 closure, grey for the post-collapse era.

```{r fig2-panels, fig.width=16, fig.height=11}
shade_layer <- geom_rect(data = shade_df,
                         aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
                         inherit.aes = FALSE, fill = "grey85", alpha = 0.45)

## ---- a  Catch by fishery type and gear -------------------------------------
fg <- catchdec %>%
  filter(area %in% OCC_AREAS, Year %in% 1951:2025) %>%
  mutate(cat = case_when(
    fishery == "Reduction"                        ~ "Reduction",
    fishery == "Roe" & gear_grp == "Gillnet"      ~ "Roe - Gillnet",
    fishery == "Roe" & gear_grp == "Seine"        ~ "Roe - Seine",
    fishery == "Roe"                              ~ "Roe - Seine",
    TRUE                                          ~ "Food and bait")) %>%
  group_by(Year, cat) %>%
  summarise(catch_kt = sum(catch_t, na.rm = TRUE)/1000, .groups = "drop") %>%
  mutate(cat = factor(cat, levels = names(fishery_cols)))

p2a <- ggplot(fg, aes(Year, catch_kt, fill = cat)) +
  shade_layer + band_closure() +
  geom_col(width = 0.9) +
  scale_fill_manual(values = fishery_cols, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = "Year", y = "Catch (×1000 tonnes)") +
  common_panel_legend(pos = c(0.99, 0.99), just = c(1, 1)) +
  guides(fill = guide_legend(nrow = 2))

## ---- b  Proportion old by sampling source ----------------------------------
p2b <- ggplot(gd, aes(Year, prop_old_n, colour = src)) +
  shade_layer + band_closure() +
  geom_line(linewidth = 0.7, alpha = 0.55) +
  geom_point(size = 1.6, alpha = 0.75) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8), se = FALSE, linewidth = 1.3) +
  scale_colour_manual(values = src_cols, name = NULL) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0.03, 0.12))) +
  labs(x = "Year", y = "Proportion age ≥ 5") +
  common_panel_legend(pos = c(0.01, 0.99), just = c(0, 1)) +
  guides(colour = guide_legend(nrow = 2, override.aes = list(size = 4, linewidth = 1.4)))

## ---- c  Length-at-age ------------------------------------------------------
age_pal <- setNames(viridisLite::viridis(length(sort(unique(la$age))),
                                         option = "C", end = 0.85),
                    paste("Age", sort(unique(la$age))))
la_p <- la %>% mutate(age_lbl = factor(paste("Age", age), levels = names(age_pal)))

## Age 8 is recorded in 15 years spanning 1958-2001 with a 22-year gap, and age 9
## in a single year. A loess fitted to those interpolates straight across the gap
## and produces a trend that is an artefact of the gap, not of growth. Trend lines
## are therefore drawn only for age classes with a long, reasonably continuous
## run; the raw series is kept for every age class.
age_cover <- la_p %>% group_by(age_lbl) %>%
  summarise(n_years = n_distinct(Year),
            max_gap = max(diff(sort(unique(Year))), na.rm = TRUE), .groups = "drop")
well_sampled <- age_cover %>% filter(n_years >= 30, max_gap <= 15) %>% pull(age_lbl)
cat("Trend lines drawn for:", paste(well_sampled, collapse = ", "), "\n")

p2c <- ggplot(la_p, aes(Year, mean_len_mm/10, colour = age_lbl)) +
  shade_layer + band_closure() +
  geom_line(aes(linetype = data_source), linewidth = 0.5, alpha = 0.6) +
  geom_point(size = 1.2, alpha = 0.6) +
  geom_smooth(data = filter(la_p, age_lbl %in% well_sampled),
              aes(group = age_lbl), method = "loess", span = 0.5,
              se = FALSE, linewidth = 1.1) +
  scale_colour_manual(values = age_pal, name = "Age class") +
  scale_linetype_manual(values = c("Reduction fishery" = "dashed",
                                   "Test fishery" = "solid"), name = NULL, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(x = "Year", y = "Mean length (cm)") +
  common_panel_legend(pos = "bottom", just = "center") +
  guides(colour = guide_legend(nrow = 1, override.aes = list(size = 4, linewidth = 1.4)))

## ---- d  Proportion old in the test fishery ---------------------------------
## The statistic reported in the text is the three-phase contrast, not a trend,
## so the phase means are drawn directly rather than a smooth that averages the
## collapse dip away.
phase_means <- ta %>% group_by(phase) %>%
  summarise(m = mean(prop_age5), x0 = min(Year), x1 = max(Year), .groups = "drop")

p2d <- ggplot(ta, aes(Year, prop_age5)) +
  shade_layer + band_closure() +
  geom_line(colour = "grey35", linewidth = 0.7) +
  geom_point(aes(size = n_fish), colour = col_old, alpha = 0.75) +
  geom_segment(data = phase_means,
               aes(x = x0, xend = x1, y = m, yend = m),
               inherit.aes = FALSE, colour = col_old, linewidth = 1.6) +
  geom_text(data = phase_means, aes(x = (x0 + x1)/2, y = m, label = sprintf("%.2f", m)),
            inherit.aes = FALSE, vjust = -0.8, size = 5.5, colour = col_old) +
  scale_size_area(name = "Fish aged", max_size = 6, labels = label_number(big.mark = ",")) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     expand = expansion(mult = c(0.05, 0.15))) +
  labs(x = "Year", y = "Proportion age ≥ 5 (test fishery)") +
  common_panel_legend(pos = c(0.99, 0.99), just = c(1, 1)) +
  guides(size = guide_legend(nrow = 1))

for (nm in c("a","b","c","d")) {
  save_fig(get(paste0("p2", nm)), paste0("Fig2", nm, "_",
    c(a="catch_by_fishery_gear", b="prop_old_by_source",
      c="length_at_age", d="prop_old_test_fishery")[[nm]]))
}

fig2 <- (p2a | p2b) / (p2c | p2d) +
  plot_annotation(tag_levels = "a") &
  theme(plot.tag = element_text(size = 26, face = "bold"))

save_fig(fig2, "Figure2_size_selective_demography", width = 18, height = 12)
fig2
```

---

## Session information

```{r session}
sessionInfo()
```
