---
title: "08 - Age composition of the roe-gillnet catch"
subtitle: "Which ages the roe gillnets took, compared with the test fishery (Fig. S2)"
author: "Sarah Ohayon, Jacob T. Dingwall, Amanda E. Bates"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")
source("00_setup.R")
library(patchwork); library(scales); library(knitr)
```

Figure 2c/2d track the **population** through the test fishery. This file tracks the
**removal**: the age composition of what the roe gillnets actually landed, year by year.
The two are complementary, and the question here is a specific one - which age classes were
present in the nets, and until when did each one last appear in quantity.

```{r data}
bio_raw <- read_csv(F_BIO, show_col_types = FALSE, name_repair = "minimal", guess_max = 100000)
names(bio_raw) <- BIO_COLS

bio <- bio_raw %>%
  mutate(age       = suppressWarnings(as.numeric(age)),
         weight_g  = suppressWarnings(as.numeric(weight_g)),
         length_mm = suppressWarnings(as.numeric(length_mm)),
         stat_area = suppressWarnings(as.numeric(stat_area)),
         Year      = recode_two_digit_year(year)) %>%
  filter(stat_area %in% BIO_AREAS, !is.na(age), age >= 1)

gn <- bio %>% filter(source == "Roe Fishery", gear == "Gillnet")
tf <- bio %>% filter(source == "Test Fishery", keep_tf(source, Year, gear))   # all months, seine, from 1977

## keep only years with an adequate sample
gn_ok <- gn %>% count(Year) %>% filter(n >= MIN_FISH) %>% pull(Year)
tf_ok <- tf %>% count(Year) %>% filter(n >= MIN_FISH) %>% pull(Year)
gn <- gn %>% filter(Year %in% gn_ok)
tf <- tf %>% filter(Year %in% tf_ok)

cat(sprintf("Roe gillnet: %s aged fish, %d years (%d-%d)\n",
            format(nrow(gn), big.mark = ","), n_distinct(gn$Year),
            min(gn$Year), max(gn$Year)))
cat(sprintf("Test fishery: %s aged fish, %d years (%d-%d)\n",
            format(nrow(tf), big.mark = ","), n_distinct(tf$Year),
            min(tf$Year), max(tf$Year)))
cat(sprintf("Age range in gillnet samples: %d-%d\n", min(gn$age), max(gn$age)))
```

---

## 1  Annual age composition of the gillnet catch

```{r composition}
AGE_MAX <- 10

comp <- function(d, label) {
  d %>% mutate(age_c = pmin(age, AGE_MAX)) %>%
    group_by(Year, age_c) %>% summarise(n = n(), .groups = "drop") %>%
    group_by(Year) %>% mutate(prop = n/sum(n), n_year = sum(n)) %>% ungroup() %>%
    mutate(source = label)
}

gn_comp <- comp(gn, "Roe gillnet")
tf_comp <- comp(tf, "Test fishery")
both    <- bind_rows(gn_comp, tf_comp)
save_tab(both, "roe_gillnet_vs_test_age_composition")

age_lab <- function(a) ifelse(a == AGE_MAX, paste0(AGE_MAX, "+"), as.character(a))
```

---

## 2  Which ages were present, and until when

```{r last-appearance}
## for each age class: first and last year it appeared at all, and the last year it
## made up at least 1% of the catch
thr <- 0.01
last_tab <- gn_comp %>% group_by(age_c) %>%
  summarise(`first year present` = min(Year),
            `last year present`  = max(Year),
            `years present`      = n(),
            `last year >= 1% of catch` =
              ifelse(any(prop >= thr), max(Year[prop >= thr]), NA_integer_),
            `mean share 1972-1983` = mean(prop[Year <= 1983]),
            `mean share 1984-2024` = mean(prop[Year >= 1984]),
            .groups = "drop") %>%
  mutate(Age = age_lab(age_c),
         across(starts_with("mean share"), ~ sprintf("%.1f%%", 100*.x))) %>%
  select(Age, everything(), -age_c)
save_tab(last_tab, "roe_gillnet_age_presence")
kable(last_tab, caption = "Presence of each age class in the roe-gillnet catch")
```

---

## 3  The upper tail: how far up the age range the nets reached

```{r upper-tail, fig.width=11, fig.height=4.5}
tail_stats <- bind_rows(gn, tf %>% mutate(source = "Test Fishery")) %>%
  mutate(src = if_else(source == "Roe Fishery", "Roe gillnet", "Test fishery")) %>%
  group_by(src, Year) %>%
  summarise(q95     = quantile(age, 0.95),
            max_age = max(age),
            prop_7plus = mean(age >= 7),
            n = n(), .groups = "drop")

```

```{r tail-summary}
tail_stats %>%
  mutate(era = case_when(Year <= 1983 ~ "1972-1983",
                         Year <= 2000 ~ "1984-2000",
                         TRUE         ~ "2001-2024")) %>%
  group_by(src, era) %>%
  summarise(`mean 95th pct age` = round(mean(q95), 2),
            `mean max age`      = round(mean(max_age), 1),
            `mean share age >= 7` = sprintf("%.1f%%", 100*mean(prop_7plus)),
            years = n(), .groups = "drop") %>%
  kable(caption = "Upper tail of the age distribution by era")
```

---

## 4  Supplementary Figure S2 - where the gear starts selecting

This is the figure that justifies the age-5 threshold. Panel a is the test fishery, which
samples the spawning aggregation each spring; panel b is the roe-gillnet catch. Both are
plotted as the proportion of the aged sample in each age class, in the Fig. 2c colours, with
a black line marking the age-4/age-5 boundary.

**Why age 5 and not age 4.** Herring mature at about age 3, so age-4 fish are already repeat
spawners; a threshold at age 4 would not isolate experience. Setting it at age 5 indexes fish
carrying more than one season of spawning experience. The gear happens to draw its line in
the same place, and the occupancy models are insensitive to the choice across thresholds from
age ≥ 4 to ≥ 7 (Fig. S4).

**What panel a is and is not.** The test fishery is the closest available index of the
spawning aggregation, and it is fishery-independent in the sense that it is not a commercial
landing. It is *not* an unfished baseline. By 1975, when it begins, the population had
already been through half a century of reduction fishing, the 1967 collapse, a four-year
closure and three years of the roe fishery. Panel a therefore shows the age structure of an
**exploited** population, not a natural one - the comparison in this figure is between what
the gear takes and what the already-truncated aggregation contains, which if anything makes
the selectivity contrast conservative.

Ages 1 and 2 are pooled into a single "≤ 2" band so that both panels sum to exactly 100%;
age 1 occurs only in the test fishery and never exceeds 6% of a sample.

```{r figS2, fig.width=13, fig.height=6}
## Ages 1-2 pooled at the bottom, 10 and above pooled at the top, so every column
## in both panels sums to exactly 1. Age 1 appears only in the test fishery.
age_bin <- function(a) pmin(pmax(a, 2), 10)

age_palette <- c("2"="#2D6A4F", "3"="#40916C", "4"="#74C69D", "5"="#B7E4C7",
                 "6"="#FFD166", "7"="#F4A261", "8"="#E76F51", "9"="#C0392B",
                 "10"="#7B241C")
age_labels  <- c("2"="≤ 2", "3"="3", "4"="4", "5"="5", "6"="6",
                 "7"="7", "8"="8", "9"="9", "10"="10+")
AGES_CMP <- 2:10

## plotmath labels so only the panel letter is bold (strips parsed with label_parsed)
panel_lab <- c("Test fishery" = 'bold("a")~~"Test fishery"',
               "Roe gillnet"  = 'bold("b")~~"Roe gillnet fishery"')

## An age class absent from a given year has no row, and geom_area then draws that
## band straight from the previous year it appeared in to the next - which stacks
## out of register with the bands that do have a point at the missing year, and
## makes the column overshoot 100%. Completing the grid with explicit zeros fixes
## it: every age class has a value in every sampled year.
comp2 <- bind_rows(gn %>% mutate(src = "Roe gillnet"),
                   tf %>% mutate(src = "Test fishery")) %>%
  mutate(age_c = age_bin(age)) %>%
  group_by(src, Year, age_c) %>% summarise(n = n(), .groups = "drop") %>%
  group_by(src) %>%
  complete(Year = full_seq(Year, 1), age_c = AGES_CMP, fill = list(n = 0)) %>%
  ungroup() %>%
  ## drop years the source did not sample at all (n = 0 across every age class)
  group_by(src, Year) %>% filter(sum(n) > 0) %>%
  mutate(prop = n/sum(n)) %>% ungroup() %>%
  mutate(age_f  = factor(age_c, levels = rev(AGES_CMP),
                         labels = unname(age_labels[as.character(rev(AGES_CMP))])),
         source = factor(panel_lab[src], levels = unname(panel_lab)))

## boundary between age 4 and age 5 = proportion aged 4 or younger
thresh <- comp2 %>% group_by(source, Year) %>%
  summarise(p_young = sum(prop[age_c <= 4]), .groups = "drop")

## 1978-1979 herring gold rush: catch peaked in 1978 (the largest roe-era catch)
## and the real price per tonne peaked in 1979 at more than three times any other
## year in the series. Marked across both panels.
GOLD <- c(1977.5, 1979.5)

figS2 <- ggplot(comp2, aes(Year, prop, fill = age_f)) +
  geom_area(linewidth = 0) +
  geom_line(data = thresh, aes(Year, p_young), inherit.aes = FALSE,
            colour = "black", linewidth = 0.7) +
  annotate("rect", xmin = GOLD[1], xmax = GOLD[2], ymin = 0, ymax = 1,
           fill = NA, colour = "black", linewidth = 0.5, linetype = "22") +
  annotate("segment", x = GOLD[1], xend = GOLD[2], y = 1.035, yend = 1.035,
           colour = "black", linewidth = 0.5) +
  annotate("text", x = mean(GOLD), y = 1.055, label = "1978-79 gold rush",
           size = 3.3, vjust = 0, fontface = "italic") +
  facet_wrap(~ source, nrow = 1, labeller = label_parsed) +
  scale_fill_manual(values = setNames(unname(age_palette[as.character(rev(AGES_CMP))]),
                                      unname(age_labels[as.character(rev(AGES_CMP))])),
                    name = "Age",
                    guide = guide_legend(reverse = TRUE, keyheight = unit(13, "pt"))) +
  ## the stack tops out at exactly 1; the headroom above it carries the gold-rush label
  scale_y_continuous(labels = percent_format(1), breaks = seq(0, 1, 0.25),
                     limits = c(0, 1.10), expand = expansion(mult = c(0, 0))) +
  ## common limits so the two panels are read against the same time axis, even
  ## though the test fishery series starts three years later than the gillnet one
  scale_x_continuous(breaks = seq(1975, 2020, 10), limits = c(1972, 2024),
                     expand = expansion(mult = c(0, 0))) +
  ## no title: the figure is described by its legend in the Supplement
  labs(x = NULL, y = "Proportion of aged sample") +
  theme_herring(12) +
  theme(strip.text = element_text(face = "plain", size = 15, hjust = 0),
        axis.text  = element_text(size = 14, colour = "black"),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 13), legend.title = element_text(size = 14),
        panel.spacing.x = unit(14, "pt"))

save_fig(figS2, "FigureS2_age_composition_population_vs_gillnet", width = 13, height = 6)
figS2
```

```{r stack-check}
## every column must sum to exactly 1 in both panels
comp2 %>% group_by(source, Year) %>% summarise(total = sum(prop), .groups = "drop") %>%
  group_by(source) %>%
  summarise(min = round(min(total), 6), max = round(max(total), 6), years = n(),
            .groups = "drop") %>%
  kable(caption = "Check: stacked proportions sum to 1 in every year of both panels")
```

### Dating the gold rush

The "herring gold rush" is marked at **1978-1979**, and the two series that define it peak in
consecutive years: catch peaks in 1978 and the real price per tonne peaks in 1979.

```{r goldrush}
annual <- read_csv(file.path(DERIVED, "annual_series.csv"), show_col_types = FALSE)
annual %>% filter(Year %in% 1972:1985) %>%
  transmute(Year,
            `Catch (kt)` = round(catch_total_t/1000, 1),
            `Real value per tonne (2024 CAD)` = round(value_per_ton)) %>%
  kable(caption = "Catch and real landed value per tonne through the gold-rush years")

gr <- annual %>% filter(!is.na(value_per_ton))
cat(sprintf("Peak real value: %s CAD/t in %d - %.1f times the next-highest year (%s CAD/t in %d)\n",
            format(round(max(gr$value_per_ton)), big.mark = ","),
            gr$Year[which.max(gr$value_per_ton)],
            max(gr$value_per_ton)/max(gr$value_per_ton[-which.max(gr$value_per_ton)]),
            format(round(max(gr$value_per_ton[-which.max(gr$value_per_ton)])), big.mark = ","),
            gr$Year[order(-gr$value_per_ton)][2]))
cat(sprintf("Peak roe-era catch: %.1f kt in %d\n",
            max(annual$catch_total_t[annual$Year >= 1972], na.rm = TRUE)/1000,
            annual$Year[annual$Year >= 1972][which.max(annual$catch_total_t[annual$Year >= 1972])]))
cat(sprintf("1978 value was %.1f times the 1977 value; 1979 was %.1f times 1978\n",
            gr$value_per_ton[gr$Year == 1978]/gr$value_per_ton[gr$Year == 1977],
            gr$value_per_ton[gr$Year == 1979]/gr$value_per_ton[gr$Year == 1978]))
```

```{r threshold-numbers}
thresh %>%
  mutate(era = if_else(Year <= 1983, "1975-1983", "1984-2024")) %>%
  group_by(source, era) %>%
  summarise(`mean proportion age >= 5` = sprintf("%.0f%%", 100*mean(1 - p_young)),
            years = n(), .groups = "drop") %>%
  kable(caption = "Proportion of the sample aged 5 or older, by source and era")
```

```{r crossover-stats}
## the quantitative claim behind the figure and the caption
cross <- both %>% filter(age_c %in% 2:10) %>%
  select(Year, age_c, source, prop) %>%
  pivot_wider(names_from = source, values_from = prop) %>%
  filter(!is.na(`Roe gillnet`), !is.na(`Test fishery`), `Test fishery` > 0) %>%
  mutate(ratio = `Roe gillnet`/`Test fishery`,
         era   = if_else(Year <= 1983, "1975-1983", "1984-2024")) %>%
  group_by(Age = age_c) %>%
  summarise(`median ratio` = round(median(ratio), 2),
            `years gillnet > test` = sprintf("%d / %d", sum(ratio > 1), n()),
            `median ratio 1975-1983` = round(median(ratio[era == "1975-1983"]), 2),
            `median ratio 1984-2024` = round(median(ratio[era == "1984-2024"]), 2),
            .groups = "drop")
save_tab(cross, "gillnet_selectivity_ratio_by_age")
kable(cross, caption = paste("Over-representation of each age class in the roe-gillnet catch",
                             "relative to the test fishery. A ratio above 1 means the gear",
                             "takes that age class disproportionately."))
```

---

## Session information

```{r session}
sessionInfo()
```
